```{r setup}
# install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

5 main dplyr functions

-   filter(): subset rows in a dataset based on data values

-   arrange(): reorder the rows

-   select(): subset columns in a dataset based on column names

-   mutate(): create new variables with functions of existing variables

-   summarise(): collapse many values down to a single summary

We will be practicing with `dplyr` functions using the `starwars` dataset:

```{r}
head(starwars)
?starwars
```

## Subset rows with `filter`

filter allows us to keep rows that match a condition

```{r}
## return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## How to get all characters that are NOT human?
filter(starwars, species != "Human")
```

`variable_name %in% vector_of_values` returns `TRUE` if the value of `variable_name` is equal to one of the values in `vector_of_values`. This is a shortcut for `variable_name == vector_of_values[1] | variable_name == vector_of_values[2]`...

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan") # returns FALSE bc "grey" is not in vector of colors

## How would we filter for characters with brown or blond hair?
filter(starwars, hair_color %in% c("blond", "brown"))
```

`grpl("characters_of_interest", variable_name)` returns `TRUE` if the value of `variable_name` contains the `characters_of_interest`.

```{r}
filter(starwars, grepl("blond", hair_color))

## How would we get all characters with blue in their eyes?
filter(starwars, grepl("blue", eye_color))
```

You can use `<`, `>`, `<=`, `>=` to compare numeric variables or categorical variables

-   nominal categorical variables (character vectors) are ranked alphabetically

-   ordinal categorical variables (factors) are ranked based on the built-in order

```{r}
## starwars characters with a name starting with letters before "D"
filter(starwars, name < "D") # ranking is based on alphabetical order

str(diamonds)
filter(diamonds, cut > "Good") # ranking is based on levels of the ordered factor cut

"D" < "d"
"d" < "D"
```

`between(x, left, right)` is a shorthand for `x >= left & x <= right`

```{r}
between(5, 1, 7) # TRUE bc 5 is greater than 1 and less than 7
filter(starwars, between(mass, 50, 100))
```

### Practice!

Find all starwars characters that:

1.  are missing a hair color
2.  are shorter than 100cm and not droids
3.  were born between the years 100 and 200
4.  another of your choice!

```{r}
# 1. missing a hair color
filter(starwars, is.na(hair_color))

# 2. shorter than 100cm and not droids
filter(starwars, height < 100 & !grepl("Droid", species))
filter(starwars, height < 100 & species != "Droid")
filter(starwars, height < 100, species != "Droid")

# 3. born between the years 100 and 200
filter(starwars, between(birth_year, 100, 200))

# 4. humans from tattooine
filter(starwars, homeworld == "Tatooine", species == "Human")
```

## Order rows with `arrange`

`arrange()` reorders rows based on the values of column(s)

```{r}
## order by character names
arrange(starwars, name)
# order of character values: aAbBcCdD...

## order by species and birth year
arrange(starwars, species, birth_year)

## use desc() to get the birth years in descending order
arrange(starwars, species, desc(birth_year))
```

### Practice!

1.  Arrange the starwars dataset from tallest to shortest characters
2.  Arrange the starwars dataset from shortest to tallest by species
3.  Alphabetize the starwars characters by their hair color, then by their eye color

```{r}
# 1. tallest to shortest
arrange(starwars, desc(height))

# 2. shortest to tallest by species
arrange(starwars, height, species) # order by height, tiebreaker is alphbetic species
arrange(starwars, species, height) # order by species, 

# 3. alphabetical by hair color
arrange(starwars, hair_color, eye_color)
```

## Subset columns with `select`

```{r}
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can also select multiple columns in a row

```{r}
# all columns between name and eye_color (inclusive)
select(starwars, name:eye_color)
```

We can also remove columns by name using a `-`

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

### Helpful functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything()) # put species first, then everything else
select(starwars, "species", everything()) # same thing

## alphabetize columns
starwars_var_names <- sort(names(starwars))
starwars_var_names
select(starwars, starwars_var_names)
```

`starts_with("abc")` matches column names that start with "abc"

`ends_with("abc")` matches column names that end with "abc"

`contains("abc")` matches column names that contain "abc"

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))
```

### Practice!

1.  Select each character's `homeworld` and `species`. What question might this subset of variables help answer?
2.  Brainstorm as many ways as possible to select `hair_color`, `skin_color`, `eye_color`, and `birth_year`

```{r}
# 1. 
select(starwars, homeworld, species)
# we might look at this info to analyze the main homeworlds of each species

# 2. 
select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, hair_color:birth_year)
select(starwars, contains("_"))
select(starwars, contains("color"), birth_year)
select(starwars, -c(name:mass, sex:starships))
```

## Add new variables with `mutate`

```{r}
starwars_small <- select(starwars, height, mass, species) # subset to practice mutate with
starwars_small

mutate(starwars_small, 
       height_m = height/100)
```

We can also create multiple variables at once and we can refer to variables we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

## only keep new variables with transmute()
transmute(starwars_small, 
          height_m = height/100, 
          bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: `sum()`, `mean()`

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm=T))

mutate(starwars_small, 
       row_addition = height + mass, 
       summary_sum = sum(height, mass, na.rm=T))
```

Group a variable using the `.by` parameter

```{r}
mutate(starwars_small,
       species_mass = sum(mass, na.rm=T), 
       .by = species)
```

Change the type of a variable

```{r}
mutate(starwars_small, 
       mass = as.character(mass))
```

### Practice!

1.  convert mass to pounds (1 kg \~ 2.2 lbs)

```{r}
mutate(starwars_small, 
       mass_lbs = mass * 2.2)
```

## Summarize data with `summarize` (or `summarise`)

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm=T))

# Group by species
summarize(starwars, 
          mean_mass = mean(mass, na.rm=T), 
          .by = species)
```

The only columns kept are the columns you group by and the columns you create.

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars, 
          mean_mass = mean(mass, na.rm=T),
          count = n(), 
          .by = species)

## group by multiple groups
summarize(starwars, 
          mean_mass = mean(mass, na.rm=T),
          count = n(), 
          .by = c(species, sex))
```

## Renaming

`rename(data, new_name = old_name)`

```{r}
new_starwars <- rename(starwars, 
       birthYear = birth_year)

new_starwars
```
